Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2023-12-28 Manuscript_Nater_etal.qmd
Added some references

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions (e.g. + ref on basic EBV idea), yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others (+ref). This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets (+ ref). On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller (+ ref). This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space (+ref). This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes (+ref). Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>), and many more. <!-- is there any other one we want to name here specifically? -->

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref). Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

-   How variance decomposition was done

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R package "targets" [@landau2021]. A visual representation of the pipeline is displayed in @fig-workflow.

@@

@@

The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>. The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).

>>>>

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes \[@proencca2017\]. Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>), and many more. <!-- is there any other one we want to name here specifically? -->

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all other components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021]. The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc. A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

@@

All numerical results in the following are presented as median \[95% credible interval unless otherwise indicated. Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv".

@@

The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub:vv. The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).
